15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
update methods pt. 4
update authors and methods

Luca L. Morreale^1,2^*

@@

William J. McShea^1^

@@

The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (NZCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W). The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@anonymous_smithsonian_2011, @bourg_initial_2013, @mcgarvey_effects_2013a].. The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world [@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].

In addition to our consideration of the entire forest plot, we focus further on three upland forest portions of the plot (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as “Low deer, low canopy vulnerability”: a 4-ha portion of upland forest in the SE quarter of the plot (Fig. **1a**), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is “high deer, low canopy vulnerability”: 16.3 ha of upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. **1b**). The third portion of the plot is "high deer, high canopy vulnerability": corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021a. We defined vulnerability to non-indigenous insects and pathogens based on the abundance (>20% of tree basal area) of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment in 2008. The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].

@@

We estimated total aboveground biomass (AGB) from the dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* [@gonzalez-akre_allodb_2022]. The *allodb* package creates species-specific allometric equations by weighting all applicable allometries within its database filtered by tree size, geography, and taxonomic information. Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot study regions and the entire 25.6-ha plot. We quantified trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region. We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data (excluding stems < 10 cm dbh).

>>>>

Luca L. Morreale^1,2^* ( Orcid ID: 0000-0003-2023-9150))

@@

William J. McShea^1^ ( Orcid ID : 0000-0002-8102-0200)

@@

The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (NZCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W). The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@anonymous_smithsonian_2011, @bourg_initial_2013, @mcgarvey_effects_2013a]. The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world [@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].

In addition to our consideration of the entire forest plot, we focus further on three upland forest portions of the plot (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first focal area we define as “low deer, low canopy vulnerability”: a 4-ha portion of upland forest in the SE quarter of the plot (Fig. **1a**), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second portion is “high deer, low canopy vulnerability”: 16.3 ha of upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. **1b**). The third focal area of the plot is "high deer, high canopy vulnerability": corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021a. We defined vulnerability to non-indigenous insects and pathogens based on the abundance (>20% of tree basal area) of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment in 2008. The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].

@@

We estimated total aboveground biomass (AGB) from the dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* [@gonzalez-akre_allodb_2022]. The *allodb* package creates species-specific allometric equations by weighting all applicable allometries within its database filtered by tree size, geography, and taxonomic information. Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot regions, as well as the entire 25.6-ha plot. We quantified trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region. We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data (excluding stems < 10 cm dbh).
